---
title: "Analytical Plan for Effect of socioeconomic status in mortality rates after brain injury: retrospective cohort"
subtitle: 'DOCUMENT: SAP-2023-004-BH-v02'
author: '**From:** Felipe Figueiredo **To:** Brennan Hickson'
date: "2023-06-01"
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    reference_docx: misc/style_SAP_en.docx
    toc: yes
# toc-title: "Sumário"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(scipen = 999)
library(pander)
library(knitr)
panderOptions('table.style', 'rmarkdown')
panderOptions('table.split.table', 160)
```

---

\newpage

**Analytical Plan for Effect of socioeconomic status in mortality rates after brain injury: retrospective cohort**

**Document version**

```{r history}
Version <- c("01", "02")
Changes <- c("Initial version", "Added methods to account for time-dependent covariates")
history <- cbind(Version, Changes)
colnames(history) <- c("Version", "Alterations")
# pander(history, split.cells = 70)
kable(history)
```

```{r cleaning, include=FALSE}
source("scripts/input.R", encoding = 'UTF-8')
```

---

# Abbreviations

- FIM: Functional Independence Measure
- HR: hazards ratio
- SD: standard deviation
- SES: socioeconomic status

# Context

## Objectives

To determine the effect of socioeconomic status of the neighborhood on mortality of patients with brain injury.

## Hypotheses

The neighborhood to which an individual is discharged from acute care affects the mortality rates of individuals that suffered a brain injury.

# Data

## Raw data

The raw data was received in three distinct databases, one containing clinical and epidemiological data, a second one with follow up on the status of individuals and a third one containing SES information of each Zip code.
The clinical and followup tables were merged by study ID, and this was joined with the SES table by the Zip code of the neighborhood of discharge, resulting in the original raw data base.

<!-- Before joining the tables by zip code, the missing locations at discharged were assumed to not have changed since injury, and those values were used to fill missing zip data where available. -->
<!-- To maximize the availability of SES data, this was also assumed for each follow up location, where the preferred information used was the zip code at discharge, when available. -->

The original data base had
`r Nvar_orig`
variables collected on
`r Nobs_orig`
observations from
`r Nobs_orig_id` individuals.

<!-- Upon study start the raw data will be collected in a raw table, that will be processed before analysis. -->
<!-- The raw dataset to be collected will have -->
<!-- `r Nvar_orig` -->
<!-- variables collected on -->
<!-- `r Nobs_orig` -->
<!-- observations. -->

<!-- This dataset will include the dates of entry and exit of the cohort, or the date of hospital admission and the date where the endpoint was reached (either an event, or hospital discharge). -->
<!-- Table 1 shows the structure of the raw dataset. -->

<!-- ```{r analytical data} -->
<!-- analytical_mockup %>% -->
<!--   pander(caption = "**Table 1** Raw dataset structure.") -->
<!-- ``` -->

<!-- Each row represents all information collected from each study participant, and each participant included will require a unique study ID. -->
<!-- The outcome should be recorded as a binary variable: either the study participant reached the endpoint -->
<!-- ([outcome]) -->
<!-- or reached the end of study period without experiencing the event. -->
<!-- This information can be recorded in either text form (eg, yes/no), or an indicator -->
<!-- ([outcome] = 1, -->
<!-- end of observation time or discharge = 0). -->
<!-- Exposure and all comorbidities should be recorded as binary variables, whenever possible. -->
<!-- All comorbidities should be recorded as binary variables, whenever possible. -->

<!-- Surgery length should be recorded in -->
<!-- minutes, -->
<!-- hours, -->
<!-- and ICU length in -->
<!-- days. -->
<!-- weeks. -->
<!-- months. -->

<!-- Surgery type will be recorded as a categorical variable. -->
<!-- It is recommended that the number of classes (types of surgery) be as small as possible, to avoid loss of precision in the analysis estimates, without the need to increase the sample size. -->
<!-- Broader categories of surgeries should be preferred, whenever possible. -->

## Analytical dataset

The date of death of participants was available, and its presence was used as an indicator of the living status of the participant at each follow up.
All variables in the raw dataset had varying missing data codes in the data dictionary made available by the researcher, which were used to attribute missingness status to each datum.
Most of the categorical variables were measured with many levels that were condensed into fewer levels for analysis.

<!-- After the cleaning process -->
<!-- `r Nvar_final` -->
<!-- variables were included in the analysis -->
<!-- with -->
<!-- `r Nobs_final` -->
<!-- observations. -->
After the cleaning process
`r Nvar_final`
variables were included in the analysis.
The total number of observations excluded due to incompleteness and exclusion criteria will be reported in the analysis.
<!-- Table 1 shows the structure of the analytical dataset. -->

```{r analytical data}
# analytical_mockup %>%
#   pander(caption = "**Table 1** Analytical dataset structure after variable selection and cleaning.")
```

All variables in the analytical set were labeled according to the raw data provided and values were labeled according to the data dictionary for the preparation of production-quality results tables and figures.

# Study parameters

## Study design

Retrospective cohort.

## Inclusion and exclusion criteria

**Inclusion criteria**

- Participants with at most 10 years of follow up;
- Participants included in the cohort between `r study_period[1]` and `r study_period[2]`.
- Participants with non-negative time under observation.

**Exclusion criteria**

Observations after 2019-12-31 will be excluded in order to mitigate risk of confounding by COVID-19 related deaths.
Observations prior to this date will still be considered for participants where such data is available.

## Exposures

SES of the neighborhood to which the participant was discharged.
The SES measure was stratified into its quintiles, and labelled according to the data dictionary to facilitate interpretation of the results.

## Outcomes

**Specification of outcome measures** (Zarin, 2011):

1. (Domain) Mortality
2. (Specific measurement) Death
3. (Specific metric) Time-to-event
4. (Method of aggregation) Hazard ratio

**Primary outcome**

Death after a brain injury.

## Covariates

Covariates will be included in the models in sets (section 5.1.3).

Demographic covariates:

- Sex
- Race
- Age at injury
- Education
- Employment status

Clinical covariates:

- Cause of injury
- Substance Problem Use
- Previous seizure disorder diagnosis
- Spinal cord injury
- Primary rehabilitation payer
- Days From Injury to Rehab Discharge
- FIM Motor at Discharge quartiles
- FIM Cognitive at Discharge quartiles

Geographical covariates:

- Rural area
- Residence after rehab discharge

# Statistical methods

## Statistical analyses

### Descriptive analyses

The epidemiological profile of the study participants will be described.
Demographic
and clinical variables
will be described as
mean (SD)
or as counts and proportions (%), as appropriate.
The distributions of participants' characteristics will be summarized in tables and visualized in exploratory plots.

### Inferential analyses

All inferential analyses will be performed in the statistical models (described in the next section).

### Statistical modeling

The hazard of mortality will be assessed with multivariate Cox regression models.
In order to assess if there is an effect of the SES of the neighborhood to which the participant was discharged on mortality will be assessed with four nested models.

1. A crude estimate of the HR between each SES quintiles and mortality will be calculated as the basis of interpretation of the effect.
1. Model 2 will include demographic covariates on top of the crude estimate.
1. Model 3 will add clinical covariates on top of model 2.
1. Model 4 will add geographical covariates to model 3, i.e., it will be calculated adjusting for all covariates described in section 4.5.

The estimates on the association between SES quintiles and mortality will be compared across models in a sensitivity analysis to gauge the cumulative effect of each set of covariates.

The proportional hazards assumption will be tested by assessing the the distribution of Schoenfeld residuals against time, the Schoenfeld test and the distribution of Martingale residuals of time-dependent covariates against the null model.
In order to keep the maximum number of time-dependent covariates in the final model, we plan to perform the residual analysis combining the following approaches: stratification by categorical variables, time splitting and non-linear fitting of covariates using splines.
Time-dependent covariates that fail these approaches will be removed before the final model is evaluated and described.

<!-- Finally, the same model will be fitted to a filtered dataset that excludes all deaths happening within one year, to assess if there is an effect on late mortality in a sensitivity analysis. -->

### Missing data

No missing data imputation will be performed.
All evaluations will be performed as complete case analyses.
Missing data counts and proportions will be reported in tables.

## Significance and Confidence Intervals

All analyses will be performed using the significance level of 5%.
All significance hypothesis tests and confidence intervals computed will be
two-tailed.
<!-- left-tailed. -->
<!-- right-tailed. -->

## Study size and Power

N/A

## Statistical packages

This analysis will be performed using statistical software `R` version `r getRversion()`.

# Observations and limitations

**Recommended reporting guideline**

The adoption of the EQUATOR network (<http://www.equator-network.org/>) reporting guidelines have seen increasing adoption by scientific journals.
All observational studies are recommended to be reported following the STROBE guideline (von Elm et al, 2014).

<!-- In particular when a retrospective study is conducted using hospital records, it is recommended that the RECORD extension of the STROBE guideline is considered (Benchimol et al, 2015). -->

# References

- **SAR-2023-004-BH-v02** -- Effect of socioeconomic status in mortality rates after brain injury: retrospective cohort
<!-- - Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd Ed.). New York: Routledge. -->
- Zarin DA, et al. The ClinicalTrials.gov results database -- update and key issues. N Engl J Med 2011;364:852-60 (<https://doi.org/10.1056/NEJMsa1012065>).
- Gamble C, et al. Guidelines for the Content of Statistical Analysis Plans in Clinical Trials. JAMA. 2017;318(23):2337–2343 (<https://doi.org/10.1001/jama.2017.18556>).
- von Elm E, Altman DG, Egger M, Pocock SJ, Gøtzsche PC, Vandenbroucke JP; STROBE Initiative. The Strengthening the Reporting of Observational Studies in Epidemiology (STROBE) Statement: guidelines for reporting observational studies. Int J Surg. 2014 Dec;12(12):1495-9 (<https://doi.org/10.1016/j.ijsu.2014.07.013>).

# Appendix

This document was elaborated following recommendations on the structure for Statistical Analysis Plans (Gamble, 2017) for better transparency and clarity.

## Availability

All documents from this consultation were included in the consultant's Portfolio.

<!-- The client has requested that this analysis be kept confidential until a future date, determined by the client. -->
<!-- All documents from this consultation are therefore not published online and only the title and year of the analysis will be included in the consultant's Portfolio. -->
<!-- After the agreed date is reached, the documents will be released. -->

<!-- The client has requested that this analysis be kept confidential. -->
<!-- All documents from this consultation are therefore not published online and only the title and year of the analysis will be included in the consultant's Portfolio. -->

The portfolio is available at:

<https://philsf-biostat.github.io/SAR-2023-004-BH/>

## Associated analyses

This analysis is part of a larger project and is supported by other analyses, linked below.

**Migration patterns in a cohort of individuals with brain injury in the US**

<https://philsf-biostat.github.io/SAR-2023-016-BH/>

**Sensitivity of mortality rates to imputation of missing socioeconomic data after brain injury: cohort study**

<https://philsf-biostat.github.io/SAR-2023-017-BH/>
